Cross Validation Approaches for Penalized Cox Regression

### 2019-05-24 main.tex
Submitted to Statistics in Medicine

\begin{figure}
  \centering
  \includegraphics[height= 12cm ]{./manuscript_figure/figure_1_new.png}
  \caption{\label{Fig:methods} Illustration of $K$-fold cross validation methods for Cox regression.}
\end{figure}

@@

The fundamental challenge of conducting cross validation for Cox regression is that the baseline hazard is not estimated from the model. Hence, another approach to cross-validation would be to include an extra step of estimating the baseline hazard and using it to quantify the cross-validation error. The basic approach to estimating the cumulative baseline hazard, $\Lambda_0$, is the Nelson-Aalen Estimator \citep{nelson1969, aalen1978}:
\begin{equation}
  \hat{\Lambda}_{0}(t) = \sum_{t_j \leq t}\frac{\text{number of failures at time } t_j}{\text{number at risk right before time }t_j}.
\end{equation}
The Cox model implies the following relationship between the baseline cumulative hazard and the cumulative hazard for individual $i$:
\begin{equation}
  \hat{\Lambda}_{i}(t) =  \hat{\Lambda}_{0}(t)\exp(X_i\Tr\hat\beta).
\end{equation}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[height= 7cm ]{./manuscript_figure/figure_2_new.png}
  \caption{\label{Fig:mse-brier-c} Simulation comparing 4 CV methods. The horizontal axis in all three plots is $s$, which represents the signal strength in the data. The value of $s$ is varied from 0.4 to 0.9.  $\log(\text{MSE ratio})$, relative to the oracle modle is plotted in the left panel. Out-of-sample Brier scores are plotted in the middle. Out-of-sample C index is plotted in the right panel. For each simulated data set, n = 120, p = 1000. Expected censoring percentage is 10$\%$.}
\end{figure}

@@

\begin{figure}[!htb]
  \centering
  \includegraphics[height= 7cm ]{./manuscript_figure/figure_3.png}
  \caption{\label{Fig:stability} Simulations illustrating the stability issues with the basic approach. For the figure on the left, $n = 100$, fold assignments are unbalanced, and the vertical axis is the proportion of undefined partial likelihoods. For the figure on the right, $n = 120$, fold assignments are balanced so that the number of events in each fold is the same, and the vertical axis is the basic deviation of the squared error loss. Basic cross-validation is much more variability than other approaches when there are only few events in each fold.}
\end{figure}

@@

The results of this simulation are illustrated in the right panel of Figure~\ref{Fig:stability}. While the other three approaches are not affected by the number of folds and number of events in each fold, the variability of basic cross-validation increases dramatically as the number of events per fold decreases. When there are sufficient number of events in each fold, the basic approach is similar to the other three approaches. However, when there are only one or two events in each fold, the basic approach can be up to 10 times more variable than other methods. Thus, although the approach can be used to calculate CVE, there is considerable uncertainty surrounding its choice of the best model. The same phenomenon is also observed for the mean squared error. When there are only few events in each fold, MSE of the basic approach is affected and increases with the increasing uncertainty.

@@

Based on the simulation results illustrated in this section, we suggest that one should proceed with caution if they want to use the basic approach: i) always balance the censoring status when conducting fold assignments for the basic approach, ii) avoid using the basic approach in conjuction with high number of folds, iii) avoid using the basic approach when there are few number of events observed. Overall we would recommend using the linear predictor approach than the basic approach in practice.

@@

\begin{figure}[!htb]
  \centering
  \includegraphics[height= 8cm ]{./manuscript_figure/figure_4.png}
  \caption{\label{Fig:loocv}Simulation comparing LOO and 10-fold CV across three different approaches. With $N = 200$ replications, the data sets generated have dimension $n = 120$ and $p = 1000$. The vertical axis is the $\log$(MSE ratio). In each panel, the LOO version and 10-fold version of the same approach are plotted. In all three panels, 10-fold basic cross-validation is included as a baseline for comparison (blue line).}
\end{figure}

For this simulation experiment, we generated datasets with $n = 120$ and $p = 1000$, using the same data generating procedure as described in Section~\ref{Sec:accuracy}. For each of the V\&VH, linear predictor, and deviance residual approaches, we conducted both 10-fold CV and LOOCV. We also conducted the basic 10-fold cross-validation as a baseline for comparison. We recorded the model that minimizes the cross-validated error  for each approach. After $N = 200$ replications, $\log$ of the MSE ratio was computed using the same definition described in Section~\ref{Sec:accuracy}.

@@

\begin{figure}[!htb]
  \centering
  \includegraphics[height= 15cm ]{./manuscript_figure/figure_auc.png}
  \caption{\label{Fig:cvauc}Comparing Partial Likelihood and C Index both built from Cross-validated Linear Predictors. The horizontal axis represents the strength of signal $s$ in the data. The left figures are under the low dimension scenarios where p = 50 and n = 150. The right figures are under the high dimension scenarios where p = 1000 and n = 120. The expected percentages of censoring for both scenarios are 30 $\%$. The top row are box plots of the selected tuning parameter$\lambda$. The mid row plots the $\log(\text{relative Squared Error Loss})$, with respect to the oracle model. The bottom row plots the Brier Scores. }
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[height= 8cm ]{./manuscript_figure/shedden.png}
  \caption{\label{fig:CVE}Rescaled Cross Validated Errors at $\alpha = 0.3$. The horizonal axis is $\log(\lambda)$. The vertical axis is the cross-validated error rescaled to a range from 0 to 1 in order to allow plotting of the four methods on a common axis. The vertical dashed lines represent the $\lambda$ selected by a specific method. The linear predictor approach selected the smallest $\lambda$ and Verweij and Van Houwelingen's approach selected the greatest $\lambda$.}
\end{figure}

@@

Cross-validation is a crutial step in fitting penalized Cox regression models. In this paper, we propose two cross-validation methods for Cox Regression: cross-validated linear predictors and cross-validated deviance residuals. We conducted a series of simulation experiments to study and compare their performance with two existing approaches, the basic cross-validation and Verweij and Van Houwelingen's cross-validated partial likelihood, in selecting tuning parameters when fitting penalized regression models.

>>>>

\begin{figure}
  \centering
  \includegraphics[height= 12cm ]{figure_1_new.png}
  \caption{\label{Fig:methods} Illustration of $K$-fold cross validation methods for Cox regression.}
\end{figure}

@@

The fundamental challenge of conducting cross validation for Cox regression is that the baseline hazard is not estimated from the model. Hence, another approach to cross-validation would be to include an extra step of estimating the baseline hazard and using it to quantify the cross-validation error. The standard approach to estimating the cumulative baseline hazard, $\Lambda_0$, is the Nelson-Aalen Estimator \citep{nelson1969, aalen1978}:
\begin{equation}
  \hat{\Lambda}_{0}(t) = \sum_{t_j \leq t}\frac{\text{number of failures at time } t_j}{\text{number at risk right before time }t_j}.
\end{equation}
The Cox model implies the following relationship between the baseline cumulative hazard and the cumulative hazard for individual $i$:
\begin{equation}
  \hat{\Lambda}_{i}(t) =  \hat{\Lambda}_{0}(t)\exp(X_i\Tr\hat\beta).
\end{equation}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[height= 7cm ]{figure_2_new.png}
  \caption{\label{Fig:mse-brier-c} Simulation comparing 4 CV methods. The horizontal axis in all three plots is $s$, which represents the signal strength in the data. The value of $s$ is varied from 0.4 to 0.9.  $\log(\text{MSE ratio})$, relative to the oracle modle is plotted in the left panel. Out-of-sample Brier scores are plotted in the middle. Out-of-sample C index is plotted in the right panel. For each simulated data set, n = 120, p = 1000. Expected censoring percentage is 10$\%$.}
\end{figure}

@@

\begin{figure}[!htb]
  \centering
  \includegraphics[height= 7cm ]{figure_3.png}
  \caption{\label{Fig:stability} Simulations illustrating the stability issues with the basic approach. For the figure on the left, $n = 100$, fold assignments are unbalanced, and the vertical axis is the proportion of undefined partial likelihoods. For the figure on the right, $n = 120$, fold assignments are balanced so that the number of events in each fold is the same, and the vertical axis is the standard deviation of the squared error loss. Basic cross-validation is much more variable than other approaches when the number of events in each fold is small.}
\end{figure}

@@

The results of this simulation are illustrated in the right panel of Figure~\ref{Fig:stability}. While the other three approaches are not affected by the number of folds and number of events in each fold, the variability of basic cross-validation increases dramatically as the number of events per fold decreases. When there is a sufficient number of events in each fold, the basic approach is similar to the other three approaches. However, when there are only one or two events in each fold, the basic approach can be up to 10 times more variable than other methods, with a corresponding increase in the mean squared error.  In these situations, therefore, even though the basic approach is defined and CVE can be calculated, its performance is very poor.

@@

Based on the simulation results illustrated in this section, we suggest that one should proceed with caution if they want to use the basic approach: (i) it is best to balance the censoring status when creating fold assignments for the basic approach, (ii) avoid using the basic approach in conjunction with a high number of folds, and (iii) avoid using the basic approach when the number of observed events is small. Alternatively, use the linear predictor approach, which has similar performance without the stability issues.

@@

\begin{figure}[!htb]
  \centering
  \includegraphics[height= 8cm ]{figure_4.png}
  \caption{\label{Fig:loocv}Simulation comparing LOO and 10-fold CV across three different approaches. With $N = 200$ replications, the data sets generated have dimension $n = 120$ and $p = 1000$. The vertical axis is the $\log$(MSE ratio). In each panel, the LOO version and 10-fold version of the same approach are plotted. In all three panels, 10-fold basic cross-validation is included as a baseline for comparison (blue line).}
\end{figure}

For this simulation experiment, we generated datasets with $n = 120$ and $p = 1000$, using the same data generating procedure as described in Section~\ref{Sec:accuracy}. For each of the V\&VH, linear predictor, and deviance residual approaches, we conducted both 10-fold CV and LOOCV. We also conducted basic 10-fold cross-validation as a baseline for comparison. We recorded the model that minimizes the cross-validated error  for each approach. After $N = 200$ replications, $\log$ of the MSE ratio was computed using the same definition described in Section~\ref{Sec:accuracy}.

@@

\begin{figure}[!htb]
  \centering
  \includegraphics[height= 15cm ]{figure_auc.png}
  \caption{\label{Fig:cvauc}Comparing Partial Likelihood and C Index both built from Cross-validated Linear Predictors. The horizontal axis represents the strength of signal $s$ in the data. The left figures are under the low dimension scenarios where p = 50 and n = 150. The right figures are under the high dimension scenarios where p = 1000 and n = 120. The expected percentages of censoring for both scenarios are 30 $\%$. The top row are box plots of the selected tuning parameter $\lambda$. The mid row plots the $\log(\text{relative Squared Error Loss})$, with respect to the oracle model. The bottom row plots the Brier Scores. }
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[height= 8cm ]{shedden.png}
  \caption{\label{fig:CVE}Rescaled Cross Validated Errors at $\alpha = 0.3$. The horizonal axis is $\log(\lambda)$. The vertical axis is the cross-validated error rescaled to a range from 0 to 1 in order to allow plotting of the four methods on a common axis. The vertical dashed lines represent the $\lambda$ selected by a specific method. The linear predictor approach selected the smallest $\lambda$ and Verweij and Van Houwelingen's approach selected the greatest $\lambda$.}
\end{figure}

@@

Cross-validation is a crutial step in fitting penalized Cox regression models. In this paper, we propose two cross-validation methods for Cox Regression: cross-validated linear predictors and cross-validated deviance residuals. We conducted a series of simulation experiments to study and compare their performance with two existing approaches, basic cross-validation and Verweij and Van Houwelingen's cross-validated partial likelihood, in selecting tuning parameters when fitting penalized regression models.

### 2019-05-24 sub/sim/coxcv.tex
Submitted to Statistics in Medicine

\documentclass{article}
\usepackage{graphicx}
\usepackage{color}

\usepackage{alltt}
\usepackage{amssymb,amsmath,graphicx,enumerate,subcaption,tikz,url,booktabs}
\usepackage[bmargin=0.75in, tmargin =0.75in,lmargin = 0.5in,rmargin = 0.5in]{geometry}
\graphicspath{{manuscript_figure/}}

\usepackage[numbers]{natbib}
\usepackage{bbold}

@@

\title{Cross Validation Approaches for Penalized Cox Regression}
\author{Biyue Dai\\Department of Biostatistics\\University of Iowa
  \and
  Patrick Breheny\\Department of Biostatistics\\University of Iowa}
\date{\today}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\maketitle

\section*{Abstract}
Cross validation is commonly used for selecting tuning parameters in penalized regression, but its use in penalized Cox regression models has received relatively little attention in the literature. Due to its partial likelihood construction, carrying out cross validation for Cox models is not straightforward, and there are several potential approaches for implementation. Here, we propose two new cross-validation methods for Cox regression and compare them to approaches that have been proposed elsewhere.
Our proposed approach of cross-validating the linear predictors seems to offer an attractive balance of performance and numerical stability. We illustrate these advantages using simulated data as well as using them to analyze data from a high-dimensional study of survival in lung cancer patients.
\input{main}

%\bibliographystyle{abbrvnat}
\bibliographystyle{unsrtnat}
%\setcitestyle{round} 
\bibliography{articles}

>>>>

\documentclass[AMA,STIX1COL]{WileyNJD-v2}
\articletype{Research Article}
\raggedbottom

\usepackage{color}
\usepackage{alltt}
\usepackage{amssymb,amsmath,graphicx,enumerate,subcaption,booktabs}
%\usepackage[bmargin=0.75in, tmargin =0.75in,lmargin = 0.5in,rmargin = 0.5in]{geometry}
\graphicspath{{../../manuscript_figure/}}
\usepackage{natbib}
\usepackage{bbold}
\hypersetup{colorlinks=true, citecolor=black, urlcolor=cyan}

@@

\begin{document}

\title{Cross Validation Approaches for Penalized Cox Regression}

\author[1]{Biyue Dai}

\author[1]{Patrick Breheny}

\authormark{Dai and Breheny}

\address[1]{\orgdiv{Department of Biostatistics}, \orgname{University of Iowa}, \orgaddress{\state{Iowa}, \country{USA}}}

\corres{Patrick Breheny \\ Department of Biostatistics \\ 145 N Riverside Dr, N336 CPHB \\ Iowa City, IA 52242 \\ \email{patrick-breheny@uiowa.edu}}

\abstract[Summary]{Cross validation is commonly used for selecting tuning parameters in penalized regression, but its use in penalized Cox regression models has received relatively little attention in the literature. Due to its partial likelihood construction, carrying out cross validation for Cox models is not straightforward, and there are several potential approaches for implementation. Here, we propose two new cross-validation methods for Cox regression and compare them to approaches that have been proposed elsewhere.
Our proposed approach of cross-validating the linear predictors seems to offer an attractive balance of performance and numerical stability. We illustrate these advantages using simulated data as well as using them to analyze data from a high-dimensional study of survival in lung cancer patients.}

\keywords{Model selection, Cross validation, Cox regression, Lasso, Penalized regression}

\jnlcitation{\cname{%
\author{Dai B} and
\author{Breheny P}} (\cyear{2019}),
\ctitle{Cross Validation Approaches for Penalized Cox Regression}, \cjournal{Statistics in Medicine}, \cvol{xxxx; xx:xx-xx}.}

\maketitle

\input{main}

\subsection*{Data and code availability}

Code to reproduce all results in this manuscript, including the raw data for the gene expression analysis, is available at \url{https://github.com/biydai/Cross-Validation}.  NOTE: This repository is currently private, but will be made public upon acceptance of this manuscript.

\subsection*{Supporting information}

A short document supporting our approach for estimating $\Lambda_0$ in the deviance residual method is provided as supporting information.

\bibliographystyle{WileyNJD-AMA}
\bibliography{../../articles}
